---
title: "Mapping Telemetry Observations"
---

```{r}
#| include: false
akbs_locs <- readRDS(here::here("shared_cache","akbs_locs_filt.rds"))
```

## Mapping in R

There are a number of workflows within R for making maps based on geospatial
data. Here, we focus almost exclusively on mapping within the `ggplot2`
package ecosystem. But, it is certainly worth the effort to learn about the
other options including `tmap`, `mapview`, and `leaflet`.

## Choosing a Projection

It's important that your spatial coordinates be projected. This is an important
step when presenting your data as a map because the choice of projection can
lead to unintentional perception bias and cause confusion. Just as importantly,
geographic data that is represented in latitude and longitude coordinates cannot
be used for animal movement modeling. Ideally, you will already have a good
understanding of your study area and will have determined your target CRS. In
this situation, you can use the `sf::st_transform()` function to transform your
coordinates into this new projection.

If you are unsure of what CRS might be a good choice for your data, the
`crsuggest` package can provide some insight. We'll use this package to
find a sensible projection for our Alaska bearded seal data. To help narrow
the options, we can specify that we want to keep the corresponding geographic
coordinate system as WGS 84 by passing the `4326` EPSG code to the `gcs`
argument. We can also specify that we want our units to be in meters.

```{r}
#| warning: false
#| message: false
library(sf)
library(crsuggest)

top_crs <- suggest_crs(akbs_locs,
            gcs = 4326,
            units = "m")

top_crs %>% 
  dplyr::select(crs_code, crs_name, crs_gcs, crs_units)
```
 The first CRS suggested is the `r top_crs$crs_name[1]`. So, let's go ahead
 and transform our data and see what this projection looks like on a map with
 our data. As mentioned above, we'll use the `sf::st_transform()` function
 to do the transformation. To give ourselves some context, we'll load simple
 world basemap data from the `rnaturalearth` package.
 
```{r}
#|: message: false
#|: warning: false
#| fig-asp: 1
#| fig-cap: Basic map of locations from bio-loggers deployed on six
#|   bearded seals in northwest Alaska
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)

world <- ne_countries(scale = "medium",returnclass = "sf") %>% 
  sf::st_make_valid()
crs_code <- as.integer(top_crs$crs_code[1])
 
filt_locs <- akbs_locs %>% sf::st_transform(crs_code)
world <- world %>% sf::st_transform(crs_code) 

map <- ggplot() + geom_sf(data = world) +
  geom_sf(data = filt_locs, size = 0.5, shape = 19) + 
  coord_sf(
    xlim = sf::st_bbox(filt_locs)[c(1,3)],
    ylim = sf::st_bbox(filt_locs)[c(2,4)]) +
  theme_minimal()

map
```
 
This looks pretty good and, now, we can proceed with making a more complete
set of maps for our observations. The focus is still more about understanding
the nature and distribution of our data than making perfect maps. But, it
is reasonable to expect these products might be useful within future
publications or presentations.

## Map Observations with `ggplot2`

There are three methods for visualizing our observation data:

1. points on a map, colored by `deploy_id`
2. lines on a map, colored by `deploy_id`
3. grid of point density

### Points on a Map

For this, we can start with the figure we made above. But, we'll want to be more
deliberate about grouping, color choices, and labels. For color choices, we're
going to choose palettes from the `colorspace` package. Also, we'll take the
opportunity to reduce the length of our `deploy_id` values.

```{r}
#| warning: false
#| message: false
#| fig-asp: 1
#| fig-cap: Map of locations from bio-loggers deployed on six
#|   bearded seals in northwest Alaska
library(colorspace)
library(stringr)

filt_locs <- filt_locs %>% 
  dplyr::mutate(deploy_id = stringr::str_sub(deploy_id, 1,11))

map <- ggplot() + geom_sf(data = world) +
  geom_sf(data = filt_locs, aes(color = deploy_id),
          size = 0.35, shape = 19) + 
  coord_sf(
    xlim = sf::st_bbox(filt_locs)[c(1,3)],
    ylim = sf::st_bbox(filt_locs)[c(2,4)]) +
  scale_color_discrete_qualitative(
    palette = "Dark 3",
    name = "deployment") +
  labs(title = "Observed Locations of 6 Bearded Seals in NW Alaska",
        subtitle = paste0("some data were censored based on a",
                          "speed, distance, and angle filter")) +
  theme_minimal()

map

```

This is pretty good, but it would be nice if we could interact with the data and
zoom in on particular regions. The `mapview` package provides an easy means for
creating interactive maps of your spatial data. The one caveat to `mapview` is
that it is based on the Web Mercator projection (a la Google Maps, etc) and
there are some extra steps required if your tracks cross the dateline at 180
(which ours do ... yay! fun!)

What we need to do is transform our data back to geographic and, then,
convert the coordinates from -180:180 to 0:360. We will use a custom
written function to handle this for us. The `st_to_360()` function
is not needed for data sets that do not cross the 180 meridian.

```{r}
#| label: st-to-360
#| warning: false
#| message: false
st_to_360 <- function(g) {
  coords <- (sf::st_geometry(g) + c(360,90)) %% c(360) - c(0,90)
  g <- sf::st_set_geometry(g,coords) %>% sf::st_set_crs(4326)
  return(g)
}
```

We will use the ESRI Ocean Basemap for our background layer and the same
qualitative color palette we used above. The `mapview` package is based on the
web mapping library, `leaflet`. Because of that, these interactive maps are
only available when rendering to HTML or other HTML capable viewers. 

```{r wrap-dateline}
#| warning: false
#| message: false
#| label: filt-locs-mapview
#| fig-cap: Interactive map of locations from bio-loggers deployed on six
#|   bearded seals in northwest Alaska
library(mapview)

sf::st_transform(filt_locs,4326) %>% st_to_360() %>% 
mapview::mapview(map.types = "Esri.OceanBasemap",
                 zcol = "deploy_id",
                 layer.name = "Deployment ID",
                 col.regions = qualitative_hcl(palette = "Dark 3", n =6),
                 cex = 1.5, lwd = 0)
```

### Lines on a Map

In actuality, the point observations from our deployments represent an 
underlying movement path for each animal. And, later, we'll work to create
a movement model that estimates this path. But, for now, we can approximate
this by simply connecting the points into an ordered path. The `sf` package
can represent a range of spatial data types, including _LINESTRING_. Since we have
multiple lines within our data set, we'll create a _MULTILINESTRING_. Creating
a spatial line from a set of spatial points isn't, initially, intuitive. We need
to ensure key attributes are retained during the conversion. First, we need to
identify our 'grouping' variable. In this case, as with our points, the 
`deploy_id` identifies our groups. Within each group, our points need to be
properly ordered. The `date` column provides this order for us. Lastly, when
we combine the individual points into a line, all of the point level attributes
(e.g. date, location type, error) will be lost. If there are key attributes
that should be retained, then those need to either be included as grouping
variables or stored in a table that can be, later, joined with the lines
data. We are going to keep things simple and not worry about line attributes
beyond `deploy_id`.

```{r}
#| warning: false
#| message: false
filt_lines <- filt_locs %>% 
  dplyr::group_by(deploy_id) %>% 
  dplyr::arrange(date) %>% 
  dplyr::summarise(do_union = FALSE) %>% 
  sf::st_cast("MULTILINESTRING")
```

We can use much of the same code as before and just replace `filt_locs` with
`filt_lines`.

```{r}
#| warning: false
#| message: false
#| fig-asp: 1
#| fig-cap: Map of connected movement paths from bio-loggers deployed on six
#|   bearded seals in northwest Alaska
library(colorspace)
library(stringr)

map <- ggplot() + geom_sf(data = world) +
  geom_sf(data = filt_lines, aes(color = deploy_id),
          lwd = 0.5) + 
  coord_sf(
    xlim = sf::st_bbox(filt_lines)[c(1,3)],
    ylim = sf::st_bbox(filt_lines)[c(2,4)]) +
  scale_color_discrete_qualitative(
    palette = "Dark 3",
    name = "deployment") +
  labs(title = "Connected Movement Paths of 6 Bearded Seals in NW Alaska",
        subtitle = paste0("some data were censored based on a",
                          "speed, distance, and angle filter")) +
  theme_minimal()

map

```

And, just as before, we can create an interactive map with `mapview`, this
time, showing our connected movement paths. Note that, before, we relied on the
`col.regions` parameter to specify colors for our points. For lines, we specify
the color palette with `color`. Also, note, the `lwd` was increased to **1.5**
and the `cex` specification was removed.

```{r wrap-dateline}
#| warning: false
#| message: false
#| label: filt-lines-mapview
#| fig-cap: Interactive map of connected movement paths from bio-loggers deployed on six
#|   bearded seals in northwest Alaska
library(mapview)

sf::st_transform(filt_lines,4326) %>% st_to_360() %>% 
mapview::mapview(map.types = "Esri.OceanBasemap",
                 zcol = "deploy_id",
                 layer.name = "Deployment ID",
                 color = qualitative_hcl(palette = "Dark 3", n =6),
                 lwd = 1.5)
```


### Point Density Map

Over-plotting occurs when multiple points are plotted on top of one another 
because of their close proximity. This can make it difficult fully understand
the distribution of points. To address this, we can create a uniformspatial grid 
and
summarize the number of points within each grid cell. This density grid should
provide a more representative visualization of our data.

The first step in this process is to create a spatial grid tesselation. We need
to decide on a cell resolution and the shape of our grid cells. There are no
specific rules regarding grid cell resolution. Although, the memory and
computational time required to process grid data increases quickly as changes in
cell dimensions lead to non-linear increases in the number of cells. A spatial
grid can be any regular tesselation of the study area. We have three possible
regular tesselations: equilateral triangles, squares, and regular hexagons. Of
those, squares are the most common but hexagons have some key benefits. 

:::{.callout-note}
### Why Hexagons
The following explanation comes from  Matt Strimas-Mackey at 
https://strimas.com/post/hexagonal-grids/

Regular hexagons are the closest shape to a circle that can be used for the
regular tessellation of a plane and they have additional symmetries compared to
squares. These properties give rise to the following benefits.

- Reduced edge effects: a hexagonal grid gives the lowest perimeter to area ratio of any regular tessellation of the plane. In practice, this means that edge effects are minimized when working with hexagonal grids. This is essentially the same reason beehives are built from hexagonal honeycomb: it is the arrangement that minimizes the amount of material used to create a lattice of cells with a given volume.
- All neighbours are identical: square grids have two classes of neighbours, those in the cardinal directions that share an edge and those in diagonal directions that share a vertex. In contrast, a hexagonal grid cell has six identical neighbouring cells, each sharing one of the six equal length sides. Furthermore, the distance between centroids is the same for all neighbours.
- Better fit to curved surfaces: when dealing with large areas, where the curvature of the earth becomes important, hexagons are better able to fit this curvature than squares. This is why soccer balls are constructed of hexagonal panels.
:::

Creating our grid within R is pretty straightforward using the 
`sf::st_make_grid()` function. We'll set our cellsize
to 50,000 meters (25 km) and, then, create an index for each of our cells. The
`sf::st_join()` function will intersect the hexagonal grid with our locations
so we can count the number of locations within each cell. One thing we need to
decide is whether to aggregated the counts across all observations or keep the
grouped approach and determine separate counts for each `deploy_id`. Since we
are still interested in understanding the distribution and nature of our data
as best as possible, it seems sensible to keep the count separated by 
`deploy_id`. However, with more than 6 deployments, this could become
untenable to plot and fully explore.

```{r}
#| warning: false
#| message: false
hexgrid <- sf::st_make_grid(st_bbox(filt_locs) %>% st_as_sfc(), 
                            cellsize = 50*1000,
                            what = "polygons", square = FALSE) 
hexgrid <- sf::st_sf(index = 1:length(lengths(hexgrid)), hexgrid)

hexbin <- sf::st_join(filt_locs, hexgrid, join = st_intersects)

locs_count <- hexgrid %>%
  dplyr::left_join(
    dplyr::count(hexbin, index, deploy_id) %>%
      tibble::as_tibble() %>%
      dplyr::select(deploy_id, index, ct=n)
  ) %>% tidyr::drop_na()
```


```{r}
#| warning: false
#| message: false
#| fig-asp: 1
#| fig-cap: Maps showing the spatial distribution of location observations 
#|   from bio-loggers deployed on six
#|   bearded seals in northwest Alaska
library(scico)
map <- ggplot() + geom_sf(data = world) +
  geom_sf(data = locs_count, 
          size = 0.125,
          aes(fill = ct)) +
  coord_sf(
    xlim = sf::st_bbox(filt_lines)[c(1,3)],
    ylim = sf::st_bbox(filt_lines)[c(2,4)]) +
  scale_fill_scico(palette = 'imola',
                    trans = "log10",
    name = "number of locations",
    guide = guide_colorbar(title.position = "bottom", barwidth = 15, 
                           barheight = 0.5, title.hjust = 0.5)) + 
  facet_wrap(~ deploy_id) +
  theme(legend.position = "bottom") +
  ggtitle("Spatial distribution of bearded seal bio-logger locations") +
  labs(caption = bquote("One hexagonal cell represents 50"~km^2))
  
map
```

