---
title: "Package Installation & Dependencies"
execute:
  eval: false
---

## Update Your Version of R

The examples and workflows described here are all developed with **R version 4.2**
and you are strongly encouraged to update your version to 4.2 _or greater_. In
general, the code and examples provided should work reliably with any version
of R greater than 4.0.

## Install Core Packages

The focus of this site, and the example workflows we describe, is on animal
movement modeling in R using the `{crawl}` package. In addition, the recently
developed `{pathroutr}` and `{crawlUtils}` packages are key enhancements that
extend the capabilities of `{crawl}` and improve the user experience.

You are strongly encouraged to install the latest available versions of each
package.

### Install `{crawl}`

#### Install via CRAN

`{crawl}` is currently available on CRAN and R >= 4.0 is highly recommended. 

```{r}
# install latest version of crawl from CRAN
install.packages("crawl")
```


#### Install via R-Universe

The latest version of `{crawl}` is also available via R-Universe.

```{r}
# Install crawl from my R-Universe repository
# Enable repository from dsjohnson
options(repos = c(
  dsjohnson = 'https://dsjohnson.r-universe.dev',
  CRAN = 'https://cloud.r-project.org'))
# Download and install crawl in R
install.packages('crawl')
# Browse the crawl manual pages
help(package = 'crawl')
```

You can also add the repository to your local list of repositories in your
*.Rprofile* and this will ensure `update.packages()` pulls any new releases
of `{crawl}` from R-Universe

```{r}
#install.packages("usethis")
usethis::edit_r_profile()
# add the following text or replace existing repos option
options(repos = c(dsjohnson = 'https://dsjohnson.r-universe.dev',
                  CRAN = 'https://cloud.r-project.org'))
```

#### Install via Github

A development version  of `{pathroutr}` is also available from
[GitHub](https://github.com/NMML/crawl). This version should be used with 
caution and only after consulting with package authors.

```{r}
# install.packages("remotes")
remotes::install_github("NMML/crawl@devel")
```


### Install `{pathroutr}`

`{pathroutr}` is currently not available on CRAN and also requires R >= 4.0. Please 
upgrade your version of R, if needed, before proceeding. Future versions of 
`{pathroutr}` may support pre-4.0 versions of R. But, for now, only 4.0+ is supported.

#### Install via R-Universe

Starting with v0.2.1, `{pathroutr}` is available via R-Universe.

```{r}
# Install new pathroutr version from my R-Universe repository
install.packages("pathroutr", repos = "https://jmlondon.r-universe.dev")
```

You can also add my repository to your local list of repositories in your
*.Rprofile* and this will ensure `update.packages()` pulls any new releases
of `{pathroutr}`

```{r}
#install.packages("usethis")
usethis::edit_r_profile()
# add the following text or replace existing repos option
options(repos = c(jmlondon = 'https://jmlondon.r-universe.dev',
                  CRAN = 'https://cloud.r-project.org'))
```

#### Install via Github

The development version  of `{pathroutr}` is available from
[GitHub](https://github.com/jmlondon/pathroutr) with:

```{r}
# install.packages("remotes")
remotes::install_github("jmlondon/pathroutr")
```


### Install `{crawlUtils}`

#### Install via R-Universe

```{r}
# Install crawlUtils from Devin's R-Universe repository
options(repos = c(
  dsjohnson = 'https://dsjohnson.r-universe.dev',
  CRAN = 'https://cloud.r-project.org'))
# Download and install crawlUtils in R
install.packages('crawlUtils')
# Browse the crawlUtils manual pages
help(package = 'crawlUtils')
```

You can also add the repository to your local list of repositories in your
*.Rprofile* and this will ensure `update.packages()` pulls any new releases
of `{crawlUtils}` from R-Universe

```{r}
#install.packages("usethis")
usethis::edit_r_profile()
# add the following text or replace existing repos option
options(repos = c(dsjohnson = 'https://dsjohnson.r-universe.dev',
                  CRAN = 'https://cloud.r-project.org'))
```

#### Install via Github

The development version  of `{crawlUtils}` is available from
[GitHub](https://github.com/dsjohnson/crawlUtils) with:

```{r}
# install.packages("remotes")
remotes::install_github("dsjohnson/crawlUtils")
```
