---
title: "Deployment Summary Tables"
---

## Understanding Our Source Data

Now that we've setup a data structure for efficient access, imported the
source data into R, and converted that data into a spatial data set, it's
time to explore and see what we have to work with. This is an important step so
you can recognize any problems with the data or inconsistencies that need to be
further investigated.

Summary tables are a good way of splitting large data into components of 
interest and learning how our data might be distributed across those
components. One example might be a simple calculation of the number of
location observations within each month across the individual animals. This
might identify anomalies such as locations in months prior to deployment or
missing data when it was expected.

```{r}
#| include: false
akbs_locs <- readRDS(here::here("shared_cache","akbs_locs.rds"))
```

We'll first want to group our location records by deployment and month.

```{r}
#| warning: false
#| message: false
library(sf)
library(lubridate)
library(dplyr)

dat <- akbs_locs %>% 
  sf::st_drop_geometry() %>% 
  mutate(month = lubridate::month(date)) %>% 
  group_by(deploy_id, month) %>% 
  count(name = "num_locs") 
```

To create a sensible table, let's just focus on a single _deploy_id_,
_EB2009_3000_06A1346_

```{r}
#| warning: false
#| message: false
library(knitr)

dat %>% 
  dplyr::filter(deploy_id == "EB2009_3000_06A1346") %>% 
  dplyr::arrange(month) %>% 
  knitr::kable()
```

One oddity that immediately stands out is the lack of locations in May.
This, however, is to be expected as these deployments started in June
and stopped transmitting in April of the following year and matches
expectations for battery life.

If we look at another deployment, _EB2009_3001_06A1332_, we can see
that this deployment ended in March.

```{r}
#| warning: false
#| message: false
dat %>% 
  dplyr::filter(deploy_id == "EB2009_3001_06A1332") %>% 
  dplyr::arrange(month) %>% 
  knitr::kable()
```

Lastly, deployment _EB2011_3001_10A0552_ appears to have stopped
transmitting in January of the following year which is 3-4 months
earlier than any of the other devices. 

```{r}
#| warning: false
#| message: false
dat %>% 
  dplyr::filter(deploy_id == "EB2011_3001_10A0552") %>% 
  dplyr::arrange(month) %>% 
  knitr::kable()
```

This is not so unusual, but such an anomaly is worth investigating
further to ensure there were no issues with the data processing or
other important deployment metadata.
